Automatically Splitting a Staged Lambda Calculus

### 2015-07-09 paper/semantics.tex

Lastly, to case on a term $e$ of type $\curr(A+B)$ at world \bbonem, we introduce the
$\caseP{e}{x.e_1}{x.e_2}$ construct, whose branches are world \bbonem\ terms
open on $\curr{A}$ and $\curr{B}$ respectively.
Importantly, this feature allows information to flow from \bbonep\ to \bbonem,
which would not be possible otherwise.

@@

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This contrast stems from differing goals: metaprogramming explicitly intends to model code writing code.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a subtitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

>>>>

These features are sufficient to ensure that mixed code does not leak into ground code,
however they also prevent information from ever escaping a $\curr$ wrapper.
So to allow the latter behavior but not the former, we introduce the $\caseP{e}{x.e_1}{x.e_2}$ construct, 
whose predicate is of type $\curr(A+B)$ and
whose branches are world \bbonem\ terms open on $\curr{A}$ and $\curr{B}$ respectively. 
This essentially allows code at \bbonem\ to inspect an injection tag within a $\curr$.

@@

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This contrast stems from differing goals: metaprogramming explicitly intends to model code manipulating code,
whereas our applications feel more natural with a eager interpretation of {\tt next}.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a subtitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}
